---
title: "exposomeSC quick start: donor-pseudobulk smoking association"
author: "Cuiwei Gao"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{exposomeSC quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# What the package estimates

`exposomeSC` estimates donor-level exposure associations with relative gene
expression within a cell type. Cells are measurements nested within a donor,
not independent replicates. The implemented default workflow aggregates raw
counts by donor and cell type and fits a robust edgeR quasi-likelihood model.

The current real-data application is ever- versus never-smoking in the HLCA.
It is an adjusted observational contrast, not a causal effect.

# Prepare a validated container

`sce` must contain a raw integer `counts` assay and `colData` columns identifying
donor and cell type. `donor_design` must be numeric, have one row per donor and
use the same donor identifiers as its row names.

```{r build}
library(exposomeSC)

# A deterministic, deliberately small demonstration object. The full
# analysis uses the pinned HLCA inputs described below, not these synthetic
# counts.
donor_ids <- paste0("D", seq_len(12L))
smoking_ever <- rep(c(0, 1), each = 6L)
cell_donor <- rep(donor_ids, each = 24L)
cell_type <- rep(
    rep(c("AT2", "Macrophage"), each = 12L),
    times = length(donor_ids)
)
gene_ids <- paste0("Gene", seq_len(12L))
cell_ids <- paste0("cell", seq_along(cell_donor))
counts <- outer(
    seq_along(gene_ids),
    seq_along(cell_ids),
    function(gene, cell) 3L + ((11L * gene + 7L * cell) %% 8L)
)
exposed_at2 <- smoking_ever[match(cell_donor, donor_ids)] == 1 &
    cell_type == "AT2"
counts[1L, exposed_at2] <- counts[1L, exposed_at2] + 4L
storage.mode(counts) <- "integer"
dimnames(counts) <- list(gene_ids, cell_ids)

sce <- SingleCellExperiment::SingleCellExperiment(
    assays = list(counts = counts),
    colData = S4Vectors::DataFrame(
        cell_id = cell_ids,
        donor_id = cell_donor,
        ann_level_3 = cell_type
    )
)
donor_design <- cbind(smoking_ever = smoking_ever)
rownames(donor_design) <- donor_ids

scee <- build_scee(
    sce,
    donor_design,
    sample_col = "donor_id"
)

methods::validObject(scee)
exposureNames(scee)
```

For a multistudy atlas, construct the study indicators and nonlinear age terms
before building the object. Never derive those terms, the cell-type universe or
the feature family from observed exposure effects.

# Fit the donor-pseudobulk model

```{r fit}
fit <- run_sc_exwas(
    scee,
    exposure = "smoking_ever",
    celltype_col = "ann_level_3",
    sample_col = "donor_id",
    min_cells = 10L,
    min_donors = 10L,
    min_group_donors = 5L,
    method = "edgeR",
    filter_method = "fixed_support",
    filter_min_cpm = 1,
    filter_min_donors = 5L,
    filter_min_total_count = 15L,
    robust = TRUE
)
```

The model retains the complete target-cell-type library size for offsets and
TMM normalisation even when the reported feature family is restricted. The
fixed-support filter depends on counts and donor support, not smoking labels.

# Inspect results without changing the family

```{r inspect}
required_columns <- c(
    "gene", "celltype", "log2FC", "se", "statistic", "pvalue",
    "padj", "padj_global", "n_donors"
)
stopifnot(all(required_columns %in% colnames(fit)))

# Sorting is descriptive; the inferential family was fixed before this step.
head(fit[order(fit$padj_global, fit$pvalue), required_columns])
```

For the default edgeR backend, `se` is deliberately numeric `NA` because the
quasi-likelihood F test does not supply a coefficient standard error.
`statistic` is the signed square root of that QL F statistic—a directional
ranking quantity, not a Wald z statistic and not a basis for a coefficient
confidence interval. Inspect `metadata(fit)$parameters[c("se_method",
"statistic_type")]` before any downstream operation that uses uncertainty.

In the full HLCA analysis, feature annotation defines a primary
host-expression family and separate receptor/residual families after fitting.
The reporting adjustment is recomputed across each complete prespecified
gene-by-cell-type family, so do not infer that family from this abbreviated
example.

**Numerical results are pending frozen outputs.** No discovery count,
calibration statistic or biological conclusion is embedded in this vignette.

# Sensitivity analyses

Reasonable analyses should examine annotation resolution, target-cell support,
age specification, missing BMI, within-study exposure overlap, sex strata,
current and former smoking separately, and leave-one-study-out fits. Donors with
repeated biological samples require a sample-pseudobulk sensitivity with donor
blocking. Influence is assessed by refitting after donor or study deletion,
not by selectively removing observations.

# Composition is a different outcome

Relative cell-type composition should be modelled jointly at donor level. A
within-type expression coefficient does not estimate absolute abundance, and a
captured-cell proportion is affected by tissue sampling, dissociation, capture,
quality control and annotation.

# Legacy decomposition interfaces

`run_decomposed_exwas()` has been removed, and `run_erd_interventional()` is a
developmental interface that is not for confirmatory interpretation. The former
direct/indirect labels and E-values do not establish a causal pathway or an
abundance contribution. A replacement estimand is under development.

# Full reproducible analysis

The exact HLCA asset, `/raw/X` layer and SHA-256 value are listed in the
package README. Reportable outputs must include matching input and code
manifests, parameters, design diagnostics, session information and output
hashes.

# Session information

```{r session-info}
sessionInfo()
```
