---
title: "exposomeSC quick start: donor-pseudobulk smoking association"
author: "Cuiwei Gao"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{exposomeSC quick start: donor-pseudobulk smoking association}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Installation

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("exposomeSC")

# Development version from GitHub:
BiocManager::install("CuiweiG/exposomeSC")
```

# Introduction

`exposomeSC` estimates donor-level exposure associations with relative gene
expression within a cell type. Cells are measurements nested within a donor,
not independent replicates. The implemented default workflow aggregates raw
counts by donor and cell type and fits a robust edgeR quasi-likelihood model.

The current real-data application is ever- versus never-smoking in the HLCA.
It is an adjusted observational contrast, not a causal effect.

# Prepare a validated container

`sce` must contain a raw integer `counts` assay and `colData` columns
identifying donor and cell type. `donor_design` must be numeric, have one
row per donor and use the same donor identifiers as its row names.

```{r build}
library(exposomeSC)

# A small simulated demonstration object. Cells within a donor share a
# donor-level offset, which is the dependence donor pseudobulk exists to
# handle; Gene1 is given a 2.5-fold increase in exposed AT2 cells. The full
# analysis uses the pinned HLCA inputs described below, not these counts.
set.seed(20260916L)
donor_ids <- paste0("D", seq_len(12L))
smoking_ever <- rep(c(0, 1), each = 6L)
cell_donor <- rep(donor_ids, each = 24L)
cell_type <- rep(
    rep(c("AT2", "Macrophage"), each = 12L),
    times = length(donor_ids)
)
gene_ids <- paste0("Gene", seq_len(12L))
cell_ids <- paste0("cell", seq_along(cell_donor))

donor_offset <- matrix(
    stats::rnorm(length(gene_ids) * length(donor_ids), sd = 0.35),
    nrow = length(gene_ids),
    dimnames = list(gene_ids, donor_ids)
)
exposed_at2 <- smoking_ever[match(cell_donor, donor_ids)] == 1 &
    cell_type == "AT2"
cell_mean <- exp(log(25) + donor_offset[, cell_donor])
cell_mean[1L, ] <- cell_mean[1L, ] * ifelse(exposed_at2, 2.5, 1)
counts <- matrix(
    stats::rnbinom(length(cell_mean), mu = as.vector(cell_mean), size = 8),
    nrow = length(gene_ids),
    dimnames = list(gene_ids, cell_ids)
)
storage.mode(counts) <- "integer"

sce <- SingleCellExperiment::SingleCellExperiment(
    assays = list(counts = counts),
    colData = S4Vectors::DataFrame(
        cell_id = cell_ids,
        donor_id = cell_donor,
        ann_level_3 = cell_type
    )
)
donor_design <- cbind(smoking_ever = smoking_ever)
rownames(donor_design) <- donor_ids

scee <- build_scee(
    sce,
    donor_design,
    sample_col = "donor_id"
)

methods::validObject(scee)
exposureVariables(scee)
```

For a multistudy atlas, construct the study indicators and nonlinear age terms
before building the object. Never derive those terms, the cell-type universe or
the feature family from observed exposure effects.

# Fit the donor-pseudobulk model

```{r fit}
fit <- run_sc_exwas(
    scee,
    exposure = "smoking_ever",
    celltype_col = "ann_level_3",
    sample_col = "donor_id",
    min_cells = 10L,
    min_donors = 10L,
    min_group_donors = 5L,
    method = "edgeR",
    filter_method = "fixed_support",
    filter_min_cpm = 1,
    filter_min_donors = 5L,
    filter_min_total_count = 15L,
    robust = TRUE
)
```

The model retains the complete target-cell-type library size for offsets and
TMM normalisation even when the reported feature family is restricted. The
fixed-support filter depends on counts and donor support, not smoking labels.

# Inspect results without changing the family

```{r inspect}
required_columns <- c(
    "gene", "celltype", "log2FC", "se", "statistic", "pvalue",
    "padj", "padj_global", "n_donors"
)
stopifnot(all(required_columns %in% colnames(fit)))

# Sorting is descriptive; the inferential family was fixed before this step.
head(fit[order(fit$padj_global, fit$pvalue), required_columns])
```

For the default edgeR backend, `se` is deliberately numeric `NA` because the
quasi-likelihood F test does not supply a coefficient standard error.
`statistic` is the signed square root of that QL F statistic—a directional
ranking quantity, not a Wald z statistic and not a basis for a coefficient
confidence interval. Inspect `metadata(fit)$parameters[c("se_method",
"statistic_type")]` before any downstream operation that uses uncertainty.

In the full HLCA analysis, feature annotation defines a primary
host-expression family and separate receptor/residual families after fitting.
The reporting adjustment is recomputed across each complete prespecified
gene-by-cell-type family, so do not infer that family from this abbreviated
example.

Gene1 in AT2 cells heads the table, which is the one effect the simulation
plants. The counts are simulated and the cohort is a dozen donors, so this
shows that the workflow recovers a planted effect and carries no biological
conclusion.

# Sensitivity analyses

Reasonable analyses should examine annotation resolution, target-cell support,
age specification, missing BMI, within-study exposure overlap, sex strata,
current and former smoking separately, and leave-one-study-out fits. Donors with
repeated biological samples require a sample-pseudobulk sensitivity with donor
blocking. Influence is assessed by refitting after donor or study deletion,
not by selectively removing observations.

# Composition is a different outcome

Relative cell-type composition should be modelled jointly at donor level. A
within-type expression coefficient does not estimate absolute abundance, and a
captured-cell proportion is affected by tissue sampling, dissociation, capture,
quality control and annotation.

# Legacy decomposition interfaces

`run_decomposed_exwas()` has been removed, and `run_erd_interventional()` is an
unexported developmental function that is not for confirmatory interpretation.
The former direct/indirect labels and E-values do not establish a causal
pathway or an abundance contribution. A replacement estimand is under
development.

# Full reproducible analysis

The exact HLCA asset, `/raw/X` layer and SHA-256 value are listed in the
package README. Reportable outputs must include matching input and code
manifests, parameters, design diagnostics, session information and output
hashes.

# Session information

```{r session-info}
sessionInfo()
```
