---
title: "Introduction to exposomeSC: a smoking-association workflow"
author: "Cuiwei Gao"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to exposomeSC: a smoking-association workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Installation

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("exposomeSC")

# Development version from GitHub:
BiocManager::install("CuiweiG/exposomeSC")
```

# Introduction

`exposomeSC` relates donor-level exposures to expression within annotated
single-cell populations. The package stores counts, cell metadata, donor
exposures and exposure metadata in a validated
`SingleCellExposomeExperiment`, then fits models at the biological-replicate
level.

The motivating analysis contrasts ever- with never-smoking in the integrated
Human Lung Cell Atlas (HLCA). It is an observational association analysis.
Smoking history is not a randomised intervention, and the available
ever-smoking indicator combines active and former exposure. We therefore avoid
causal language.

# Unit of analysis and estimand

Cells inherit their donor's exposure. Treating those cells as independent
replicates would underestimate uncertainty. For each cell type, the canonical
workflow sums raw counts within donor, producing one pseudobulk column per
eligible donor. The target is the exposure-associated rate ratio for relative
expression within that cell type.

This target differs from cell-type abundance. If smoking changes only the
number of captured cells of a type, both its gene counts and complete library
size may change together while its within-type expression rate remains stable.
Relative composition is consequently analysed as a separate multivariate
outcome.

# Required data contract

A defensible input has:

- a raw, non-negative integer count assay;
- unique cell and feature identifiers;
- a donor identifier for every cell;
- donor-constant exposure and baseline-covariate values;
- cell-type annotations fixed before exposure testing; and
- an auditable map from biological samples to donors when donors have repeated
  samples.

The rows of the donor design matrix must match donor identifiers exactly. A
categorical study variable is represented by prespecified numeric indicator
columns before the object is built.

```{r build}
library(exposomeSC)

# A small simulated object makes the vignette executable. Cells within a
# donor share a donor-level offset, and Gene1 is raised 2.5-fold in exposed
# AT2 cells. It illustrates the data contract only; it is not used for the
# HLCA results.
set.seed(20260916L)
donor_ids <- paste0("D", seq_len(12L))
smoking_ever <- rep(c(0, 1), each = 6L)
cell_donor <- rep(donor_ids, each = 24L)
cell_type <- rep(
    rep(c("AT2", "Macrophage"), each = 12L),
    times = length(donor_ids)
)
gene_ids <- paste0("Gene", seq_len(12L))
cell_ids <- paste0("cell", seq_along(cell_donor))
donor_offset <- matrix(
    stats::rnorm(length(gene_ids) * length(donor_ids), sd = 0.35),
    nrow = length(gene_ids),
    dimnames = list(gene_ids, donor_ids)
)
exposed_at2 <- smoking_ever[match(cell_donor, donor_ids)] == 1 &
    cell_type == "AT2"
cell_mean <- exp(log(25) + donor_offset[, cell_donor])
cell_mean[1L, ] <- cell_mean[1L, ] * ifelse(exposed_at2, 2.5, 1)
counts <- matrix(
    stats::rnbinom(length(cell_mean), mu = as.vector(cell_mean), size = 8),
    nrow = length(gene_ids),
    dimnames = list(gene_ids, cell_ids)
)
storage.mode(counts) <- "integer"

sce <- SingleCellExperiment::SingleCellExperiment(
    assays = list(counts = counts),
    colData = S4Vectors::DataFrame(
        cell_id = cell_ids,
        donor_id = cell_donor,
        ann_level_3 = cell_type
    )
)
age_scaled <- as.numeric(scale(c(
    45, 52, 61, 48, 57, 66, 50, 59, 68, 54, 63, 72
)))
donor_design <- cbind(
    smoking_ever = smoking_ever,
    age_scaled = age_scaled
)
rownames(donor_design) <- donor_ids

scee <- build_scee(
    sce,
    donor_design,
    sample_col = "donor_id"
)

methods::validObject(scee)
exposureVariables(scee)
```

`exposureVariables()` lists every column of the donor design, so the age term
appears alongside the exposure. Which column is the exposure, and which are
adjustments, is set in the call to `run_sc_exwas()` below.

# Canonical donor-pseudobulk model

The implemented primary backend uses robust edgeR quasi-likelihood inference.
Expression support is fixed independently of the smoking labels, and the
complete target-cell-type transcriptome remains available for library-size and
normalisation calculations after filtering. Adjustment terms are supplied
through `covariates`, as columns of the donor design; a nonlinear age term or a
multistudy indicator set is built before the object is created, never derived
from the observed exposure effects.

```{r primary-model}
fit <- run_sc_exwas(
    scee,
    exposure = "smoking_ever",
    celltype_col = "ann_level_3",
    sample_col = "donor_id",
    covariates = "age_scaled",
    min_cells = 10L,
    min_donors = 10L,
    min_group_donors = 5L,
    method = "edgeR",
    filter_genes = TRUE,
    filter_method = "fixed_support",
    filter_min_cpm = 1,
    filter_min_donors = 5L,
    filter_min_total_count = 15L,
    robust = TRUE
)

head(fit[order(fit$padj_global, fit$pvalue), ])
```

The call above is deliberately short. An analysis meant for reporting also
builds its adjustment terms, records the input and code versions it ran on,
inspects the design for rank and separation problems, and states the feature
family before the model is fitted.

# Multiplicity and output interpretation

The primary feature family comprises the annotation-defined, exposure-blind
support-filtered host-expression transcriptome. Benjamini--Hochberg adjustment
is applied across its complete realised tested gene-by-cell-type family. The
host category is externally annotated; the realised tested universe also
depends on the donor-support rule, which is fixed before fitting.
Immune-receptor and residual-annotation
features are retained as separate secondary or exploratory audits; their
adjusted values are not merged into the primary family after looking at the
results.

The returned data include the effect estimate, backend-specific test statistic,
unadjusted p-value, within-cell-type adjustment, cross-cell-type adjustment and
donor-support diagnostics. edgeR quasi-likelihood fits report the coefficient
standard-error field as numeric `NA`: `glmQLFTest()` provides a QL F test, not a
coefficient standard error. The signed square-root QL statistic preserves
direction for ranking but is neither a Wald statistic nor a basis for a
coefficient confidence interval. DESeq2 and voom-dream retain their own valid
Wald and moderated-t uncertainty semantics. A small p-value is insufficient on
its own. We also require a stable feature definition, adequate support, an
identifiable design, sensitivity to annotation and sampling choices,
empirical-null calibration, and influence assessment.

The object built above is synthetic, so the fit illustrates the interface and
carries no biological conclusion.

# Repeated samples and composition

The donor-aggregated model is the primary expression estimand. A separate
sample-pseudobulk sensitivity preserves repeated biological samples and uses
donor blocking with voom/limma while retaining the primary hypothesis universe.
It is not a second discovery screen.

The composition analysis operates on the supported `ann_level_3` simplex. It
uses Jeffreys-regularised log ratios, donor-level aggregation and a
study-adjusted multivariate test. The resulting coefficients concern relative
captured-cell composition—not absolute lung-cell abundance—and no mediation
interpretation is made.

# Developmental decomposition interfaces

`run_decomposed_exwas()` has been removed. `run_erd_interventional()` is kept
as an unexported developmental function and is not for confirmatory
interpretation. The earlier within-type-rate outcome cannot identify the
mechanical contribution of cell-type abundance to a tissue mixture, so former
direct/indirect labels and ERD E-values must not be used for causal or abundance
claims.

A replacement estimand remains developmental until it is implemented, calibrated
under appropriate data-generating mechanisms and assessed independently.

# Public source and reproducibility

The HLCA source is the CC BY 4.0 CZ CELLxGENE collection associated with
Sikkema *et al.* (2023):

- collection: <https://cellxgene.cziscience.com/collections/6f6d381a-7701-4781-935c-db10d30de293>;
- exact H5AD: <https://datasets.cellxgene.cziscience.com/688185ad-11c2-4172-a53a-f4f1f4076860.h5ad>;
- publication: <https://doi.org/10.1038/s41591-023-02327-2>;
- count layer: `/raw/X`; and
- H5AD SHA-256: `1cbdde1e513a31bdb6a3f70296e2d0612760d2af4032e239c94596ea87b44007`.

Every reportable result must retain the verified input hashes, ordered code
manifest, model parameters, design diagnostics, session information and output
hashes. Figures and tables should read the stored result objects rather than
have numerical values transcribed into display code.

# Session information

```{r session-info}
sessionInfo()
```
