---
title: "Introduction to exposomeSC: a smoking-association workflow"
author: "Cuiwei Gao"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Introduction to exposomeSC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Scope

`exposomeSC` relates donor-level exposures to expression within annotated
single-cell populations. The package stores counts, cell metadata, donor
exposures and exposure metadata in a validated
`SingleCellExposomeExperiment`, then fits models at the biological-replicate
level.

The motivating analysis contrasts ever- with never-smoking in the integrated
Human Lung Cell Atlas (HLCA). It is an observational association analysis.
Smoking history is not a randomised intervention, and the available
ever-smoking indicator combines active and former exposure. We therefore avoid
causal language.

# Unit of analysis and estimand

Cells inherit their donor's exposure. Treating those cells as independent
replicates would underestimate uncertainty. For each cell type, the canonical
workflow sums raw counts within donor, producing one pseudobulk column per
eligible donor. The target is the exposure-associated rate ratio for relative
expression within that cell type.

This target differs from cell-type abundance. If smoking changes only the
number of captured cells of a type, both its gene counts and complete library
size may change together while its within-type expression rate remains stable.
Relative composition is consequently analysed as a separate multivariate
outcome.

# Required data contract

A defensible input has:

- a raw, non-negative integer count assay;
- unique cell and feature identifiers;
- a donor identifier for every cell;
- donor-constant exposure and baseline-covariate values;
- cell-type annotations fixed before exposure testing; and
- an auditable map from biological samples to donors when donors have repeated
  samples.

The rows of the donor design matrix must match donor identifiers exactly. A
categorical study variable is represented by prespecified numeric indicator
columns before the object is built.

```{r build}
library(exposomeSC)

# A small deterministic object makes the vignette executable. It illustrates
# the data contract only; it is not used for the HLCA results.
donor_ids <- paste0("D", seq_len(12L))
smoking_ever <- rep(c(0, 1), each = 6L)
cell_donor <- rep(donor_ids, each = 24L)
cell_type <- rep(
    rep(c("AT2", "Macrophage"), each = 12L),
    times = length(donor_ids)
)
gene_ids <- paste0("Gene", seq_len(12L))
cell_ids <- paste0("cell", seq_along(cell_donor))
counts <- outer(
    seq_along(gene_ids),
    seq_along(cell_ids),
    function(gene, cell) 3L + ((11L * gene + 7L * cell) %% 8L)
)
exposed_at2 <- smoking_ever[match(cell_donor, donor_ids)] == 1 &
    cell_type == "AT2"
counts[1L, exposed_at2] <- counts[1L, exposed_at2] + 4L
storage.mode(counts) <- "integer"
dimnames(counts) <- list(gene_ids, cell_ids)

sce <- SingleCellExperiment::SingleCellExperiment(
    assays = list(counts = counts),
    colData = S4Vectors::DataFrame(
        cell_id = cell_ids,
        donor_id = cell_donor,
        ann_level_3 = cell_type
    )
)
age_scaled <- as.numeric(scale(c(
    45, 52, 61, 48, 57, 66, 50, 59, 68, 54, 63, 72
)))
donor_design <- cbind(
    smoking_ever = smoking_ever,
    age_scaled = age_scaled
)
rownames(donor_design) <- donor_ids

scee <- build_scee(
    sce,
    donor_design,
    sample_col = "donor_id"
)

methods::validObject(scee)
exposureNames(scee)
```

# Canonical donor-pseudobulk model

The implemented primary backend uses robust edgeR quasi-likelihood inference.
Expression support is fixed independently of the smoking labels, and the
complete target-cell-type transcriptome remains available for library-size and
normalisation calculations after filtering. The HLCA analysis adjusts for a
three-degree-of-freedom natural spline in age, sex and source study.

```{r primary-model}
fit <- run_sc_exwas(
    scee,
    exposure = "smoking_ever",
    celltype_col = "ann_level_3",
    sample_col = "donor_id",
    covariates = "age_scaled",
    min_cells = 10L,
    min_donors = 10L,
    min_group_donors = 5L,
    method = "edgeR",
    filter_genes = TRUE,
    filter_method = "fixed_support",
    filter_min_cpm = 1,
    filter_min_donors = 5L,
    filter_min_total_count = 15L,
    robust = TRUE
)

head(fit[order(fit$padj_global, fit$pvalue), ])
```

The exact analysis is not reconstructed from this abbreviated call. A complete
analysis also builds the spline and study terms, verifies immutable inputs,
classifies features, checks the design, records code identity and writes
outputs atomically.

# Multiplicity and output interpretation

The primary feature family comprises the annotation-defined, exposure-blind
support-filtered host-expression transcriptome. Benjamini--Hochberg adjustment
is applied across its complete realised tested gene-by-cell-type family. The
host category is externally annotated; the realised tested universe also
depends on the frozen donor-support rule in the HLCA data. Immune-receptor and residual-annotation
features are retained as separate secondary or exploratory audits; their
adjusted values are not merged into the primary family after looking at the
results.

The returned data include the effect estimate, backend-specific test statistic,
unadjusted p-value, within-cell-type adjustment, cross-cell-type adjustment and
donor-support diagnostics. edgeR quasi-likelihood fits report the coefficient
standard-error field as numeric `NA`: `glmQLFTest()` provides a QL F test, not a
coefficient standard error. The signed square-root QL statistic preserves
direction for ranking but is neither a Wald statistic nor a basis for a
coefficient confidence interval. DESeq2 and voom-dream retain their own valid
Wald and moderated-t uncertainty semantics. A small p-value is insufficient on
its own. We also require a stable feature definition, adequate support, an
identifiable design, sensitivity to annotation and sampling choices,
empirical-null calibration, and influence assessment.

**Result values are pending frozen outputs.** This vignette contains no
discovery count, effect estimate or biological conclusion.

# Repeated samples and composition

The donor-aggregated model is the primary expression estimand. A separate
sample-pseudobulk sensitivity preserves repeated biological samples and uses
donor blocking with voom/limma while retaining the primary hypothesis universe.
It is not a second discovery screen.

The composition analysis operates on the supported `ann_level_3` simplex. It
uses Jeffreys-regularised log ratios, donor-level aggregation and a
study-adjusted multivariate test. The resulting coefficients concern relative
captured-cell composition—not absolute lung-cell abundance—and no mediation
interpretation is made.

# Developmental decomposition interfaces

`run_decomposed_exwas()` has been removed. `run_erd_interventional()` is
retained as a developmental interface and is not for confirmatory
interpretation. The earlier within-type-rate outcome cannot identify the
mechanical contribution of cell-type abundance to a tissue mixture, so former
direct/indirect labels and ERD E-values must not be used for causal or abundance
claims.

A replacement estimand remains developmental until it is implemented, calibrated
under appropriate data-generating mechanisms and assessed independently.

# Public source and reproducibility

The HLCA source is the CC BY 4.0 CZ CELLxGENE collection associated with
Sikkema *et al.* (2023):

- collection: <https://cellxgene.cziscience.com/collections/6f6d381a-7701-4781-935c-db10d30de293>;
- exact H5AD: <https://datasets.cellxgene.cziscience.com/688185ad-11c2-4172-a53a-f4f1f4076860.h5ad>;
- publication: <https://doi.org/10.1038/s41591-023-02327-2>;
- count layer: `/raw/X`; and
- H5AD SHA-256: `1cbdde1e513a31bdb6a3f70296e2d0612760d2af4032e239c94596ea87b44007`.

Every reportable result must retain the verified input hashes, ordered code
manifest, model parameters, design diagnostics, session information and output
hashes. Figures and tables read frozen result objects; numerical values are not
transcribed into display code.

# Session information

```{r session-info}
sessionInfo()
```
