---
title: "Exploratory cross-omic network inference with exposomeSC"
author: "Cuiwei Gao"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Exploratory cross-omic network inference with exposomeSC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE,
    fig.width = 7,
    fig.height = 4.5
)
```

# Installation

```{r install, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("exposomeSC")

# Development version from GitHub:
BiocManager::install("CuiweiG/exposomeSC")
```

# Introduction

The network module estimates cell-type-resolved conditional-association
structures that combine transcript pseudobulk with donor-level molecular
measurements such as metabolomics. It is exploratory. An edge is a
regularised conditional association under the fitted model—not a biochemical
interaction or a causal relation.

Network analysis answers a different question from sc-ExWAS. sc-ExWAS tests a
prespecified marginal exposure coefficient for each feature and cell type.
Network analysis describes multivariate dependence after preprocessing and
covariate adjustment. Selection and tuning uncertainty, sample size and
external stability must be reported separately.

The fitted network object records its matrices, feature annotations, method
and provenance-related metadata in an explicit S4 contract.

```{r class-contract}
library(exposomeSC)
methods::getSlots("CelltypeNetworkResult")
```

# Input alignment

The transcript and metabolite data must refer to the same donors. Raw
single-cell counts are aggregated within donor and cell type before network
estimation; donor-level metabolite rows are then aligned by identifier. Do not
match rows by position.

The examples below use a small simulated data set so that the vignette runs in
seconds: 24 donors, monocytes and B cells, 12 genes and 4 donor-level
metabolites. In monocytes four genes move with the exposure, and one metabolite
tracks it.

```{r example-data}
has_coglasso <- requireNamespace("coglasso", quietly = TRUE)
has_plot_packages <- all(vapply(c("ggplot2", "igraph", "ggraph", "tidygraph"),
    requireNamespace, logical(1), quietly = TRUE))

set.seed(2026)
donors <- sprintf("D%02d", 1:24)
celltypes <- c("Monocyte", "B cell")
cell_donor <- rep(rep(donors, each = 20), times = length(celltypes))
cell_type <- rep(celltypes, each = length(donors) * 20)
pm25 <- stats::setNames(stats::rnorm(length(donors)), donors)

counts <- matrix(stats::rpois(12 * length(cell_donor), 15), nrow = 12,
    dimnames = list(paste0("G", 1:12), paste0("c", seq_along(cell_donor))))
in_mono <- cell_type == "Monocyte"
for (gene in 1:4) {
    effect <- c(0.6, 0.6, -0.6, -0.6)[gene]
    counts[gene, in_mono] <- stats::rpois(sum(in_mono),
        15 * exp(effect * pm25[cell_donor[in_mono]]))
}
sce <- SingleCellExperiment::SingleCellExperiment(
    assays = list(counts = counts),
    colData = S4Vectors::DataFrame(cell_id = colnames(counts),
        donor_id = cell_donor, cell_type = cell_type))

donor_data <- cbind(PM2.5 = pm25,
    age = round(stats::runif(length(donors), 30, 70)),
    sex = rep(0:1, length.out = length(donors)))
rownames(donor_data) <- donors
scee <- build_scee(sce, donor_data, sample_col = "donor_id")

metabolites <- matrix(stats::rnorm(length(donors) * 4), ncol = 4,
    dimnames = list(donors, paste0("M", 1:4)))
metabolites[, "M1"] <- metabolites[, "M1"] + 1.2 * pm25
```

```{r build-network, eval=has_coglasso}
mono_net <- run_celltype_network(
    scee,
    metabolites = metabolites,
    celltype = "Monocyte",
    sample_col = "donor_id",
    covariates = c("age", "sex"),
    method = "coglasso",
    stability = TRUE,
    nlambda_w = 3L,
    nlambda_b = 3L,
    rep_num = 5L
)
```

The exposure is deliberately left out of this model. The four exposure genes
and M1 share its signal, so the network is expected to connect them; any other
edge is a false positive, which 24 donors can produce even under stability
selection. Adding `exposure = "PM2.5"` regresses the exposure out of both
matrices before estimation, and that module is no longer expected: the network
then describes dependence the exposure does not explain. Which of the two is
the scientific question has to be settled before fitting.

The tuning grid and the number of XStARS subsamples are deliberately small
here; the defaults are a 15 by 15 grid and 20 subsamples. These calls do not
reproduce the HLCA primary model. In a multistudy application, study and other
prespecified design terms must be included consistently, and the
feature-selection rule must be recorded before network fitting.

# Collaborative graphical lasso

The `coglasso` backend permits different penalties for within-omic and
between-omic precision-matrix blocks. Penalty selection is based on resampling
stability. The complete tuning grid, resample identifiers, edge-selection
frequencies and final penalty must be retained; a network plot alone is not an
adequate result record.

```{r inspect-network, eval=has_coglasso && has_plot_packages}
mono_net
plot_stability_surface(mono_net)
plot_celltype_network(
    mono_net,
    layout = "fr",
    color_by = "omic_layer",
    highlight_cross_omic = TRUE
)
```

# Comparing cell types

Cell-type comparisons require aligned features, compatible preprocessing and
adequate donor support in every network. Apparent differences can result from
unequal measurement error or selection stability rather than a true change in
conditional association.

```{r compare-networks, eval=has_coglasso && has_plot_packages}
bcell_net <- run_celltype_network(
    scee,
    metabolites = metabolites,
    celltype = "B cell",
    sample_col = "donor_id",
    covariates = c("age", "sex"),
    method = "coglasso",
    stability = TRUE,
    nlambda_w = 3L,
    nlambda_b = 3L,
    rep_num = 5L
)

comparison <- run_comparative_network(
    mono_net,
    bcell_net,
    method = "descriptive"
)
comparison

plot_network_comparison(comparison)
```

This comparison is descriptive and reports no p-values. The Fisher-Z mode is
experimental: networks estimated from the same donors are not independent
samples, and penalised partial correlations do not follow the distribution the
test assumes. Edge-presence or weight differences are not evidence of
differential conditional association.

# Exposure-restricted networks

`run_exposure_network()` uses an exposure-association stage followed by network
estimation. This is a two-stage exploratory procedure. Selecting and estimating
on the same donors can exaggerate stability and effect size; nested resampling,
sample splitting or independent validation is required for confirmatory use.

```{r exposure-network}
exposure_net <- run_exposure_network(
    scee,
    metabolites,
    celltype = "Monocyte",
    exposure = "PM2.5",
    sample_col = "donor_id",
    covariates = c("age", "sex"),
    selection_method = "univariate",
    fdr_threshold = 0.05,
    network_method = "block_glasso"
)
exposure_net
```

# Longitudinal networks

Time-point-specific networks can be organised with `run_temporal_network()`.
Repeated measurements must remain linked to their donor, and an apparent
emerging or disappearing edge is descriptive unless its uncertainty has been
assessed under an appropriate longitudinal resampling scheme.

```{r temporal-network, eval=has_plot_packages}
metabolites_follow_up <- metabolites
metabolites_follow_up[, "M2"] <- metabolites_follow_up[, "M2"] + 1.5 * pm25

temporal <- run_temporal_network(
    scee_list = list(baseline = scee, follow_up = scee),
    metabolites_list = list(
        baseline = metabolites,
        follow_up = metabolites_follow_up
    ),
    celltype = "Monocyte",
    sample_col = "donor_id",
    covariates = c("age", "sex"),
    method = "block_glasso",
    stability = FALSE
)
temporal

plot_temporal_dynamics(temporal)
```

The follow-up here reuses the baseline cells with one changed metabolite, so
the example shows the mechanics of edge tracking rather than a longitudinal
design.

# Mediation terminology

`run_network_mediation()` is a model-based exploratory utility. Network
topology does not establish temporal ordering, eliminate unmeasured
confounding or identify a causal indirect effect. Results must be described as
conditional model associations unless the study design and identification
assumptions independently support stronger language.

# Composition and provenance

Building a network within a cell type removes variation in broad cell-type
mixture from the expression outcome, but it does not eliminate subtype mixture,
sampling, dissociation, capture, annotation or residual confounding. The
network model does not supply evidence for a composition-mediated pathway.

Every reported network should retain input hashes, donor inclusion, feature
definitions, preprocessing, covariate design, tuning grid, random seeds,
resampling output, software versions and the exact object used for plotting.
The examples above run on simulated data and report no result from a real
cohort.

# References

- Albanese A, Kohlen W, Behrouzi P (2024). Collaborative graphical lasso.
  arXiv:2403.18602.
- Liu H, Roeder K, Wasserman L (2010). Stability approach to regularisation
  selection (StARS). *NeurIPS* 23.
- Meinshausen N, Bühlmann P (2010). Stability selection. *Journal of the Royal
  Statistical Society: Series B* 72:417--473.

# Session information

```{r session-info}
sessionInfo()
```
