---
title: "Exploratory cross-omic network inference with exposomeSC"
author: "Cuiwei Gao"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Exploratory cross-omic network inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE,
    message = FALSE
)
```

# Scope

The network module estimates cell-type-resolved conditional-association
structures that combine transcript pseudobulk with donor-level molecular
measurements such as metabolomics. It is exploratory. An edge is a
regularised conditional association under the fitted model—not a biochemical
interaction or a causal relation.

Network analysis answers a different question from sc-ExWAS. sc-ExWAS tests a
prespecified marginal exposure coefficient for each feature and cell type.
Network analysis describes multivariate dependence after preprocessing and
covariate adjustment. Selection and tuning uncertainty, sample size and
external stability must be reported separately.

The fitted network object records its matrices, feature annotations, method
and provenance-related metadata in an explicit S4 contract.

```{r class-contract}
library(exposomeSC)
methods::getSlots("CelltypeNetworkResult")
```

# Input alignment

The transcript and metabolite data must refer to the same donors. Raw
single-cell counts are aggregated within donor and cell type before network
estimation; donor-level metabolite rows are then aligned by identifier. Do not
match rows by position.

```{r build-network, eval=FALSE}
library(exposomeSC)

# `scee` is a validated SingleCellExposomeExperiment.
# `metabolites` has donors in rows and metabolites in columns.
mono_net <- run_celltype_network(
    scee,
    metabolites = metabolites,
    celltype = "Monocyte",
    celltype_col = "ann_level_3",
    exposure = "smoking_ever",
    covariates = c("age", "sex"),
    method = "coglasso",
    stability = TRUE,
    top_var_genes = 100L
)
```

This abbreviated call does not reproduce the HLCA primary model. In a
multistudy application, study and other prespecified design terms must be
included consistently, and the feature-selection rule must be recorded before
network fitting.

# Collaborative graphical lasso

The `coglasso` backend permits different penalties for within-omic and
between-omic precision-matrix blocks. Penalty selection is based on resampling
stability. The complete tuning grid, resample identifiers, edge-selection
frequencies and final penalty must be retained; a network plot alone is not an
adequate result record.

```{r inspect-network, eval=FALSE}
mono_net
plot_stability_surface(mono_net)
plot_celltype_network(
    mono_net,
    layout = "fr",
    color_by = "omic_layer",
    highlight_cross_omic = TRUE
)
```

# Comparing cell types

Cell-type comparisons require aligned features, compatible preprocessing and
adequate donor support in every network. Apparent differences can result from
unequal measurement error or selection stability rather than a true change in
conditional association.

```{r compare-networks, eval=FALSE}
comparison <- run_comparative_network(
    mono_net,
    bcell_net,
    method = "descriptive"
)

plot_network_comparison(comparison)
```

This comparison is descriptive. Inferential permutation and Fisher-Z modes are
retired because the current result object does not retain the aligned
donor-level matrices needed for complete paired refitting. Edge-presence or
weight differences are not evidence of differential conditional association.

# Exposure-restricted networks

`run_exposure_network()` uses an exposure-association stage followed by network
estimation. This is a two-stage exploratory procedure. Selecting and estimating
on the same donors can exaggerate stability and effect size; nested resampling,
sample splitting or independent validation is required for confirmatory use.

```{r exposure-network, eval=FALSE}
smoking_net <- run_exposure_network(
    scee,
    metabolites,
    celltype = "Monocyte",
    exposure = "smoking_ever",
    selection_method = "univariate",
    fdr_threshold = 0.05
)
```

# Longitudinal networks

Time-point-specific networks can be organised with `run_temporal_network()`.
Repeated measurements must remain linked to their donor, and an apparent
emerging or disappearing edge is descriptive unless its uncertainty has been
assessed under an appropriate longitudinal resampling scheme.

```{r temporal-network, eval=FALSE}
temporal <- run_temporal_network(
    scee_list = list(baseline = scee_baseline, follow_up = scee_follow_up),
    metabolites_list = list(
        baseline = metabolites_baseline,
        follow_up = metabolites_follow_up
    ),
    celltype = "Monocyte",
    method = "coglasso"
)

plot_temporal_dynamics(temporal)
```

# Mediation terminology

`run_network_mediation()` is a model-based exploratory utility. Network
topology does not establish temporal ordering, eliminate unmeasured
confounding or identify a causal indirect effect. Results must be described as
conditional model associations unless the study design and identification
assumptions independently support stronger language.

# Composition and provenance

Building a network within a cell type removes variation in broad cell-type
mixture from the expression outcome, but it does not eliminate subtype mixture,
sampling, dissociation, capture, annotation or residual confounding. The
network model does not supply evidence for a composition-mediated pathway.

Every reported network should retain input hashes, donor inclusion, feature
definitions, preprocessing, covariate design, tuning grid, random seeds,
resampling output, software versions and the exact object used for plotting.
No HLCA network result is claimed here; numerical outputs remain pending a
prespecified, frozen analysis.

# References

- Albanese A, Kohlen W, Behrouzi P (2024). Collaborative graphical lasso.
  arXiv:2403.18602.
- Liu H, Roeder K, Wasserman L (2010). Stability approach to regularisation
  selection (StARS). *NeurIPS* 23.
- Meinshausen N, Bühlmann P (2010). Stability selection. *Journal of the Royal
  Statistical Society: Series B* 72:417--473.

# Session information

```{r session-info}
sessionInfo()
```
