---
title: "Simulation and calibration protocol for sc-ExWAS inference"
author: "Cuiwei Gao"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Simulation and calibration protocol for sc-ExWAS inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE
)
```

# Scope

This vignette sets out the protocol for evaluating sc-ExWAS inference by
simulation: which methods can be compared, how data must be generated, which
metrics are reported and what a frozen benchmark release has to contain. It
runs two small examples, a complete-null calibration check and a power
approximation, so that the mechanics can be seen; it reports no benchmark
result, and the package makes no quantitative calibration, power, speed or
superiority claim until a frozen release does.

# Match each method to its estimand

The primary sc-ExWAS estimand is the exposure association with relative
expression within a cell type. Differential-abundance methods target relative
composition and are not expression-model competitors. A future
molecular-contribution decomposition would target the product of composition
and within-type expression and requires a distinct simulation truth.

Comparators are evaluated only when they share:

- the same biological unit of replication;
- the same raw-count input and eligibility rules;
- the same fixed feature-by-cell-type universe;
- compatible offsets and covariates;
- the same multiplicity denominator; and
- an estimand against which their output can be scored.

A cell-level test that treats cells as independent can be included as a
negative control, but not as a plausible population-inference method.

# Real-data complete-null diagnostics

The principal calibration protocol is a fixed-design negative-binomial
parametric bootstrap. The observed exposure and nuisance design, donor and
sample support, offsets, expression-support family and feature-by-cell-type
keys remain fixed; counts are generated under a reduced model with a zero
exposure coefficient, after which the frozen edgeR pipeline is rerun. The
result is a conditional plug-in diagnostic for that exact generator and family,
not proof of strong FDR control under arbitrary mixtures of true and false null
hypotheses.

The earlier within-study smoking-label shuffle is retained only as an
observational stress audit. Smoking was not randomised, and its conditional
exchangeability is not established. It is therefore ineligible as the formal
calibration, irrespective of how thoroughly its accepted permutations are
audited.

Pilot and smoke runs estimate engineering cost only. Every formal replicate
must retain its seed, fit diagnostics, complete-family call summaries, code and
input contracts, runtime and checksum. Resume logic re-authenticates completed
files before reuse, and a failed fit cannot be recoded as no call.

## A complete-null check in miniature

The simulation below has the shape of a null replicate: donor-level
biological variation in the counts, an exposure drawn independently of them,
so that every null hypothesis is true, and the default pipeline run on the
result. Five replicates of 20 donors, 60 genes and two cell types run in a
few seconds; a formal run uses far more of each and keeps every replicate's
seed and receipt.

```{r null-simulation}
library(exposomeSC)

simulate_null <- function(seed, n_donors = 20L, n_genes = 60L,
                          cells_per_donor = 30L,
                          celltypes = c("Monocyte", "NK")) {
    set.seed(seed)
    donors <- sprintf("D%02d", seq_len(n_donors))
    cell_donor <- rep(rep(donors, each = cells_per_donor),
                      times = length(celltypes))
    cell_type <- rep(celltypes, each = n_donors * cells_per_donor)
    donor_effect <- matrix(stats::rnorm(n_genes * n_donors, sd = 0.3),
                           n_genes, dimnames = list(NULL, donors))
    mu <- 8 * exp(donor_effect[, cell_donor])
    counts <- matrix(stats::rnbinom(length(mu), mu = mu, size = 10),
        nrow = n_genes,
        dimnames = list(paste0("G", seq_len(n_genes)),
                        paste0("c", seq_along(cell_donor))))
    sce <- SingleCellExperiment::SingleCellExperiment(
        assays = list(counts = counts),
        colData = S4Vectors::DataFrame(cell_id = colnames(counts),
            donor_id = cell_donor, cell_type = cell_type))
    ## drawn independently of the counts: every null is true
    design <- cbind(exposure = stats::rnorm(n_donors),
                    age = round(stats::runif(n_donors, 30, 70)))
    rownames(design) <- donors
    build_scee(sce, design, sample_col = "donor_id")
}

null_replicate <- function(seed) {
    fit <- run_sc_exwas(simulate_null(seed), exposure = "exposure",
        covariates = "age", celltype_col = "cell_type",
        sample_col = "donor_id", min_cells = 5L)
    c(tests = nrow(fit),
      calls = sum(fit$padj_global < 0.05, na.rm = TRUE),
      raw_p_below_0.05 = mean(fit$pvalue < 0.05, na.rm = TRUE))
}

replicates <- t(vapply(1001:1005, null_replicate, numeric(3)))
replicates
```

Under the complete null the family-wise any-call indicator is
`any(calls > 0)` per replicate, and the fraction of raw p-values below 0.05
should sit near 0.05:

```{r null-summary}
c(any_call_rate = mean(replicates[, "calls"] > 0),
  mean_raw_p_below_0.05 = mean(replicates[, "raw_p_below_0.05"]))
```

Five replicates cannot bound a rejection rate; the protocol asks for
independent replicates in the hundreds, with Wilson intervals over them.

# Simulation data-generating mechanisms

Simulation must generate the hierarchy explicitly:

1. study and donor covariates;
2. donor exposure under declared overlap and confounding structures;
3. donor/sample cell-type composition;
4. donor-level expression variation and dispersion;
5. per-cell RNA content and sequencing depth;
6. raw counts conditional on the preceding quantities; and
7. annotation and dropout mechanisms, where relevant.

At minimum, the design should cross null and non-null expression effects with
balanced and imbalanced exposure, heterogeneous study effects, repeated
samples, rare cell types, unequal library sizes, influential donors and
misspecified age trends. Parameter values must be justified from held-out or
external summaries, not tuned to favour one implementation.

A pure composition scenario changes only the composition-generating process.
It must not insert the composition score directly into within-type expression.
A contextual-expression scenario may do so, but it is labelled and scored as a
different mechanism.

# Metrics

For each independent simulation replicate, report:

- false-discovery proportion, defined as zero when there are no calls;
- family-wise any-call indicator under the complete null;
- power for the prespecified non-null family;
- effect-estimate bias and empirical calibration of test rejection rates;
- estimability and explicit failure reasons;
- effect concordance on a common fitted universe; coefficient standard-error
  and interval-coverage comparisons only for backends that return valid
  coefficient standard errors (explicitly not edgeR QL); and
- elapsed time and peak memory under the same hardware allocation.

False-discovery rate is the mean of replicate-level false-discovery
proportions. Gene-by-replicate rows are correlated and must not be treated as
independent observations when calculating uncertainty. Empirical-null
rejection-rate intervals are calculated over independent simulation or
bootstrap replicates. Failed fits remain failures; they are not recoded as no
call.

## A power approximation for planning

`estimate_power()` gives the large-sample power of the pseudobulk Wald test
for one gene in one cell type at the Bonferroni per-test level, from the
donor count, the log2 effect per unit exposure, the negative-binomial
dispersion and the mean pseudobulk count. It treats the dispersion as known
and is optimistic for small donor numbers; it is a planning aid, and the
simulation above is the check.

```{r power}
estimate_power(n_donors = c(20, 40, 80, 160), effect_size = 0.5,
               n_genes = 5000, n_celltypes = 5,
               dispersion = 0.1, base_mean = 100)
```

Donors, not cells, are the independent units: more cells per donor raise the
mean count and shrink its term in the standard error, but they do not add to
`n_donors`.

# Comparator alignment

The executed expression comparisons are donor-pseudobulk edgeR,
sample-pseudobulk donor-blocked voom and a deliberately naive cell-level
negative control. `muscat`, `dreamlet`, `variancePartition`, `MAST`, `DESeq2`,
`glmmTMB` and `nebula` are absent from the frozen library and must not be named
as executed. A sample-pseudobulk negative-binomial donor-random-intercept
extension using installed `lme4` is a protocol until a fixed manifest, resource
pilot and complete-family formal run pass. Method labels must describe the
actual fitted model rather than a wrapper package. Composition methods are
assessed only against composition truth.

For a targeted model evaluation, `test_features` fixes the reported family while the
full feature matrix remains available for library sizes, normalisation and
dispersion estimation. A missing or unsupported requested feature is reported
as not estimable; the family must not shrink silently.

# Multiplicity

Filtering and feature-family definitions are fixed without exposure labels.
The primary adjusted values use the complete host-expression-by-cell-type
family. Separate receptor and residual-annotation audits do not alter that
denominator. Same-data screening followed by unadjusted decomposition or
pathway inference is not a valid confirmatory procedure.

# Reproducibility record

A frozen benchmark release must include:

- the simulation and analysis source digests;
- the complete parameter grid and random-number streams;
- every simulated dataset or an exact deterministic reconstruction recipe;
- raw method calls, warnings and failures for every replicate;
- the fixed hypothesis and comparator alignment tables;
- replicate-level metrics before aggregation;
- session information, hardware and worker allocation; and
- checksums for result, summary, table and figure objects.

Figures and tables read these objects directly. No metric is typed into display
code. Until this record is complete, the package makes no quantitative
calibration, power, speed or superiority claim.

# Developmental ERD assessment

The removed decomposition interfaces and the developmental
`run_erd_interventional()` are excluded from confirmatory benchmarking. Their
within-type-rate outcome did not identify the abundance contribution described
by the former direct/indirect labels. The redesigned molecular-contribution
estimand must first be implemented, then tested under pure abundance,
within-type expression, combined, contextual, null and misspecified scenarios.
No mediational E-value is reported without a derivation valid for that exact
multivariate stochastic-composition estimand.

# Session information

```{r session-info}
sessionInfo()
```
